---
title: "titanic-exploratory-data-analysis"
author: "shobha mourya"
date: "March 4, 2019"
output: 
  html_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploratory Data Analysis

**Analysis of Titanic Passenger Surival**

Using Titanic dataset

**Using dplyr R package**

* for data wrangling and feature engineering

* dplyr makes you much more productive and far more intuitive

* Look at the help resources
?dplyr

* dplyr: a grammar of data manipulation collection of functions that makes working with data 


##Why do feature engineering?

In most Data Science scenarios, Feature Engineering is paramount

* Feature engineering often trumps algorithms, more data, parameter tuning, etc. 

Data Scientist spend 60-80% of their time on being data janitors

* getting data, understanding data, cleaning data, feature engineering commonly 
referred to as data wrangling

Faster data wrangling == more business value!

* if you can do data wrangling faster - you can compete much
much more effective

*To win Kaggle - feature engineering is most important*


In business textual data is more than numeric data
categorical data is more common
categorical data - think of dropdown with US states discrete variables
It is important to learn to work with string data


## Types of Joins

### left_join 

* all rows of primary table and matching rows of right

### right_join 

* all rows of secondary table and only matching rows of primary table

### inner_join 

* only match rows from primary and secondary tables

### full_join 

* all rows from both tables and matching rows

### semi_join 

* rows & columns only from primary table that have matching in secondary table
* no data from second table is shown

### anti-join 

* rows & columns from primary table that don't have matches in second table


## What is Imputation? 

* It is data cleansing step of data analysis in which we deduce the missing (NA) values 
* or values that are incorrect based on the domain knowledge of the case study. 

For example, in the Titanic Passenger Survival dataset, there are many observations with missing 'Age' values and there are observations where 'Fare' is zero. 
After analysing we use generalization to assign the mean/median value based on the similarity of the missing data point wrt to other data points.

## Steps for Imputing missing values

We mostly followed the below step in our imputation process:

### Stats subset

To understand the distribution of the feature we want to impute we create a subset of statistical measures like min, max, mean, median, IQR, variance(var), 
standard deviation (sd), sum/count etc. 

### Analyse stats

In this step, we view the statistics subset to analyse what would be the best value to assign, mean or median, for imputation. If the feature distribution is not skewed, that is there is not much variation between the min/max, we can use the mean for the missing values. However, if the distribution has extreme values we can use median. 
Note. Mean is a computed value affected by min and max hence in case of skewed data median is a better generalisation than mean for missing values.

### Indicator variable
We create an indicator variable with values Y (for yes) and N (for no) to serve as a flag to indicate if the observation has imputed values for a specific variable. 
For example, variable 'AgeMissing' indicates whether age is imputed or not. 
This helps in reverse engineer if at all we need to the original value in later 
stages of analysis or while building the prediction model.

### Lookup table

After we have decided on what statistics/values to use for imputation we create 
a lookup table with the new values and the common feature by which we will join the lookup table with the original table. If we're dealing with GroupWise imputation you might have to consider adding the group column and the common columns. 
For example, 'Fare' depends on the passenger class - price of first class is bound to be more than other classes and hence mean of the first class will be different than the mean of other classes.

### Left Join the lookup subset

Join the original dataset with the lookup subset using the appropriate join columns. This step adds the new columns to the original dataset - you can crosscheck using the names command.

### Impute the values

Use the mutate dplyr command to assign the new values to the feature being imputed based on the appropriate conditions. For example, use the indicator flag and then apply the condition.

### Crosscheck 

It is important to cross check that all the imputations have been applied correctly. 
Use summary command on the imputed feature to see if there are any NAs or wrong values.
If there are errors, then revisit the code and check the conditions in mutate command.

### Remove the additional columns

After assigning the appropriate statistic/value to the missing values using 
lookup table column values, their purpose is served and they should be removed 
from the original dataset. Use select(-column) or set the column to NULL to remove the columns.


# Analyse train dataset

Load the packages

```{r}
#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("stringr")

library(dplyr)
library(ggplot2)
library(stringr)

# Load raw data
path = "C:/Users/shobha/Documents/RExploratyDataAnalysis"
train = read.csv(paste0(path,"/dataFiles/train.csv"), header=TRUE)

dim(train)

str(train)
glimpse(train)
names(train)

#View(train)
```

## Variables

### PassengerId 

It is primary key not useful for ML algorithm

###Survived 

It is our label
It is the dependent variable we are trying to predict
Other variables are predictors - we need to feature engineer on predictors

###Name 

It is string variable - scope for feature engineer

### Sex and Age 

Are imp if women and children first is to be true

### Sibsp

Sibling and Spouse 
could mean travelling with spouse or sibling 

###  Parch

Parent child 
could mean traveling with parent or child

# Analyse test dataset

```{r}
test = read.csv(paste0(path,"/dataFiles/test.csv"), header=TRUE)
dim(test)
```

It has one less column

```{r}
names(test)
```

survived column is removed because we have to predict the survived label with 
our model

In order to increase our dataset we will comibine the test and train datasets.
First lets add the column Survived with None value for test dataset

```{r}

test.survived = test %>% mutate(survived = "None")
dim(test.survived)
names(test.survived)

test.survived$survived = as.factor(test.survived$survived)
train$survived = as.factor(train$survived)

str(test.survived)
str(train)

# you can also use rbind
data.combined = train %>% union_all(test.survived)

# data.combined = rbind(train, test.survived)
# dim(data.combined)
# str(data.combined)

dim(data.combined)
```

891 + 418 = 1309

Let's start with data exploration.
We will analyse the features one by one to see if they have any predictive power.

# Combine the train and test datasets 

We will use this dataset for comparing operations without dplyr

```{r}
glimpse(data.combined)

# convert survived, pclass to factor
data.combined$survived = as.factor(data.combined$survived)
data.combined$pclass = as.factor(data.combined$pclass)

glimpse(data.combined)


# Let's see the distribution of survival rates
table(data.combined$survived)


names(data.combined)
```

# Analyse the variables for their predictive power

## Explore pclass feature

```{r}
# Distribution of passengers across classes
table(data.combined$pclass)

round(prop.table(table(data.combined$pclass))*100)
```

More than 50% passengers traveled in third class.
Our Hypothesis is that rich folks, traveling in first class, survived at a higher rate

Let's visualize pclass distribution wrt to survival for train dataset

```{r}
ggplot(train, aes(x = pclass, fill=survived)) +
  geom_bar() +
  xlab("PClass") +
  ylab("Total Count") +
  ggtitle("Survival rate wrt PClass")
```

### Interprete 

The visualization shows that proportion of perished is higher in second and 
third class as compared to first class. So pclass seems to have good predictive power.

## Explore name feature

Now, let's examine the name feature

Examine the first few names in the training data set

The names have lastname, title. firstname middlename (maiden name)

```{r}
head(as.character(train$name))

tail(as.character(train$name))

#View(head(data.combined, 20))

# How many unique names are there across both train & test?
length(unique(as.character(data.combined$name)))


length(data.combined$name)
```

1307 out of 1309, so there are two duplicate names. 

```{r}
dup.names = data.combined[(duplicated(data.combined$name) == TRUE),]

dup.names
```


### Create new feature title from name

Let's create a new feature title from the name variable by extracting
the pattern which has characters followed by .

Regex - starts with char and followed by .
pattern - one or more letter [a-zA-Z] and immediately a dot
+ --> one of more, * --> zero or more

First let's try an example


```{r}
my_name = " Shobha, Ms. Mourya Shivnarayan"

my_title = str_extract(my_name, pattern = "[a-zA-Z]+\\.")
my_title

my_name = " Shobha, Miss. Mourya Shivnarayan"
my_title = str_extract(my_name, pattern = "[a-zA-Z]+\\.")
my_title


my_name = " Shobha, . Mourya Shivnarayan"
my_title = str_extract(my_name, pattern = "[a-zA-Z]+\\.")
my_title

train = train %>%
  mutate(title = str_extract(name,"[a-zA-Z]+\\."))

table(train$title)

length(unique(train$Title))
```

There are a lot of titles with only 1 occurence, Now this is not good for ML algo
for example, if there is one Jonkheer who perished then the ML algo will 
always zero-down on that and predict that a passenger with title Jonkheer will perish this is a clear case of overfitting maybe there were other Jonkheers who did survive but we don't have that data in our training set so we can collapse titles into more close title.

Title is indicative of Sex and Age. 

First create a df with two columns -  Title and New.Title 
Which is like a lookup table to map titles, for example Capt. Col. etc should be 
mapped to Mr.


Then for both dfs match on title and join the new column New.Title

Condense titles down to small subset 
left join with lookup table is a very general
pattern for replacement in dplyr
old value, new values to map


```{r}
titles.lookup = data.frame( title = c("Mr.", "Capt.",
                                      "Col.", "Don.",
                                      "Dr.", "Jonkheer.",
                                      "Major.", "Rev.",
                                      "Sir.", "Mrs.",
                                      "Dona.", "Lady.",
                                      "Mme.", "Countess.",
                                      "Miss.","Mlle.", "Ms.",
                                      "Master."),
                            new.title=c(rep("Mr.",9),
                                        rep("Mrs.",5),
                                        rep("Miss.",3),
                                        "Master."),
                            stringsAsFactors = FALSE)

#View(titles.lookup)

# Replace Titles using lookup table join lookup df for matching titles with train df

train = train %>%
  left_join(titles.lookup, by = "title")

table(train$new.title)


# Let's replace Title with New.Title and remove New.Title

train = train %>%
  mutate(title = new.title) %>%
  select(-new.title)

table(train$title)

# Specific cases is overfit - avoid avoid avoid overfitting focus on generalization

# Alternatively you could do the above in one dplyr pipeline

train = train %>%
  left_join(titles.lookup, by = "title") %>%
  mutate(title=new.title) %>%
  select(-new.title)


table(train$title)

```

Now we have four titles. Double-check our work to see if titles are assigned correctly 
that is no male is assigned Miss or Mrs and no female Mr. or Master



```{r}
train %>%
  filter((sex == "female" & (title == "Mr." | title == "Master.")) |
           (sex == "male" & (title == "Mrs." | title == "Miss"))) 


#or

table(train$title, train$sex)


#output shows one Mr is female, that's because we assigned title Mr. to Dr.

train = train %>% 
  mutate(title = ifelse((title == "Mr." & sex == "female"), "Mrs.",title))

train %>%
  filter((sex == "female" & (title == "Mr." | title == "Master.")) |
           (sex == "male" & (title == "Mrs." | title == "Miss"))) 


table(train$title, train$sex)

```

Miss - could be unmarried female child or adult


### Analysing data.combined without using dplyr

```{r}
# Analysing combined data 
title = sapply(data.combined$name, str_extract, pattern = "[a-zA-Z]+\\.")
head(title)

#View(title)

str(title)

data.combined$title = title

glimpse(data.combined)

# How many unique titles
unique(data.combined$title)

table(data.combined$title)

```

* the most common titles are Master 61, Miss 260, Mr. 757, Mrs 197 
* Other titles are rare and single digit 
* for now we will assign the rare titles as other and visualize the distribution
wrt survived label


```{r}
common_titles = c("Mr.", "Mrs.", "Miss.", "Master.")
data.combined.test = data.combined %>% 
  mutate(title = ifelse(title %in% common_titles, title, "other"))

table(data.combined.test$title)

glimpse(data.combined.test)

data.combined.test$title = as.factor(data.combined.test$title)

str(data.combined.test$title)

data.combined = data.combined.test
rm(data.combined.test)


glimpse(data.combined)

#Since we only have survived lables for the train set, only use the first 891 rows

ggplot(data.combined[1:891,], aes(x=title, fill=survived)) +
  geom_bar() +
  ggtitle("Survival rate wrt Title")

```

### Interpretation

The survival for Mr. is staggeringly lower compared to other titles.
Let's add pclass to see if sruvival of males traveling by first class is better.

```{r}
ggplot(data.combined[1:891,], aes(x=title, fill=survived)) +
  geom_bar() +
    facet_wrap(~pclass) +
    ggtitle("Survival by Pclass and Title")

```


## Explore sex feature

Let's explore the sex variable now

```{r}
str(data.combined$sex)

table(data.combined$sex)

table(data.combined$sex, data.combined$title)

# Let's visualize

ggplot(data.combined[1:891,], aes(x = sex, fill = survived)) +
  geom_bar() +
  facet_wrap(~pclass)+
  ggtitle("Survival by Pclass and Sex")

```

### Interpretation

Looks like sex is a very strong predictor of survival.
In all the three pclasses male survival rate is low. 
Females in first and second class have very high survival rate whereas in
class 3 it is almost 50/50



## Explore age feature

Let's explore the numeric variable age



```{r}
str(data.combined$age)

summary(data.combined$age)
```

There are many missing values

Just to be thorough, take a look at survival rates broken out by sex, pclass, and age
it is a numerical variable so we will plot histogram


```{r}

ggplot(data.combined[1:861,], aes(x=age, fill=survived)) +
  geom_histogram() +
  facet_wrap(~ sex + pclass) +
  ggtitle("Survival by age and pclass")

```

### Interpretation

There is lot of Age missing so we impute Age based on title and age
average or median as replacement value or avg in some cases and median in some cases


let's see the age statistics wrt Pclass and Title the older you are the wealthier you are stats and NAs for the group


```{r}
age.stats = train %>%
  group_by(title,pclass) %>%
  summarize(age.min = min(age, na.rm = T),
            age.max = max(age, na.rm = T),
            age.mean = mean(age, na.rm = T),
            age.median = median(age, na.rm = T),
            age.na.count = sum(is.na(age)),
            age.var = var(age, na.rm = T),
            age.sd = sd(age, na.rm = T),
            age.IQR = IQR(age, na.rm = T)) %>%
  arrange(title,pclass)

#View(age.stats)

table(train$pclass)

```


Rule: When mean and median are close, it is better to use mean
because it a good summary statistic

Miss is an overloaded title, it covers both unmarried adult women as well as 
female children. So Miss is problamatic for Title as proxy for Age.
So for all titles except Miss we use mean age as proxy.

Rule
Imputation Steps
Create Tracking feature/ indicator variable- Y for NA else N
lookup table

```{r}
train$age.missing = ifelse(is.na(train$age), "Y","N")

table(train$age.missing)

#View(train)
```

For matching Pclass and Title we want to replace NA age with mean / median Age
So we create a new df with Pclass, Title, mean Age, median Age
then join original df with this df on Pclass and Title
then replace Age in original with either mean Age or median Age
and remove the mean Age and median Age


```{r}
age.lookup = age.stats %>%
  select(pclass,title,age.mean,age.median)

#View(age.lookup)


train = train %>%
  left_join(age.lookup, by=c("pclass","title")) %>%
  mutate(age = ifelse(age.missing == "Y",
                      ifelse(title == "Miss.", age.median, age.mean),
                      age) ) %>%
  select(-age.median, -age.mean)

#View(train)
```


We engineered feature title as it was proxy for gender and age

We imputed the below features

zero fares - median fares based on Pclass
missing age - mean(except Miss)/median(Miss) age based on title and Pclass

Steps for imputing or replacing features

Create Tracking feature as indicator for replacing values
- it helps to reverse engineer to original value if we need to later

Create a lookup table for new features using join
- so we need to add the matching columns to the lookup table
- left join original df with new
- mutate to do the actual replacement
- remove the lookup variables from original df

Here we'll go ahead and create a whole lot of new features

look at impute Age distribution all up 5 % to 100% step by 5 percent
20 buckets

```{r}
quantile(train$age, probs = c(0.25,0.5,0.75))

quantile(train$age, probs=seq(0.05, 1, 0.05))
```

5 % have age 5.3 or below
10 % have age 16 or below
15 % have age 18 or below
80% have age 40 - so it is skewed to younger

Survival based on age - children, adults, elderly

### Analysing combined data without dplyr, just for comparison

Let's see if our titles are good indicator of age and sex


```{r}
table(data.combined$title)

boys = data.combined %>% filter(title == "Master.")
summary(boys$age)

```

Maximum age is 14.5 years, that could be considered upper limit for age for title Master.

We know that "Miss." is more complicated, let's examine further

```{r}
misses = data.combined %>% filter(title == "Miss.")
summary(misses$age)
dim(misses)

# Get subset of misses with survived not None and age is not NA
misses.survived = misses %>% filter(!is.na(age) & survived != "None")
dim(misses.survived)

# on x-axis we have age, so histogram plot, facet wrap with categorical variables pclass
ggplot(misses.survived, aes(x = age, fill = survived)) +
  geom_histogram() + 
  facet_wrap(~pclass) + 
  ggtitle("Survival for Miss. by Age broken down by Pclass")
  
# set the binwidth to 5 and plot the histogram
ggplot(misses.survived, aes(x = age, fill = survived)) +
  geom_histogram(binwidth = 5) + 
  facet_wrap(~pclass) + 
  ggtitle("Survival for Miss. by Age broken down by Pclass")

```

Female children may have better survival rate than adult females could be a candidate for feature engineering later


```{r}
#View(misses)
head(misses)

# Let's find females traveling alone, that is without parents, siblings, spouse
misses.alone = misses %>% filter(parch == 0 & sibsp == 0)

dim(misses.alone)

summary(misses.alone$age)

length(which(misses.alone$age <= 14.5))
```

## Explore sibsp variable

Let's explore the sibsp variable using data.combined dataset


```{r}
str(data.combined$sibsp)

summary(data.combined$sibsp)

# Can we treat as a factor?
length(unique(data.combined$sibsp))

unique(data.combined$sibsp)

table(data.combined$sibsp)

data.combined$sibsp <- as.factor(data.combined$sibsp)
```

Of the five variables we have explored before: names, from which we derived title,
plcass, sex, age
title represents age and sex and title & pclass have predictive power
so let's visualize survival wrt sibsp broken down by title and pclass

x-axis has sibsp which factor so bar plot facet wraped with other
factor variables pclass and title

```{r}
ggplot(data.combined[1:891,], aes(x = sibsp, fill = survived)) +
  geom_bar() +
  facet_wrap(~pclass + title) +
  ggtitle("Survival wrt sibsp broken down by Pclass and Title")
```


## Explore parch variable

Let's explore the parch variable for data.combined dataset

```{r}
str(data.combined$parch)

length(unique(data.combined$parch))

unique(data.combined$parch)

# convert it to factor and visualize to see if it has any predictive power
data.combined$parch = as.factor(data.combined$parch)
str(data.combined$parch)
table(data.combined$parch)
```

Most passengers are traveling alone, there are 2 passengers with parch size 9

x-axis parch which is factor so bar plot broken down by categorical 
variable pclass and title

```{r}
ggplot(data.combined[1:891, ], aes(x = parch, fill = survived)) +
  geom_bar() +
  facet_wrap(~pclass + title) +
  ggtitle("Survival wrt parch broken down by pclass and title")
  
```

### New feature family size

Let's create a new feature family_size

A passenger who has parch or sbsp value non-zero implies is traveling with family
sbsp is for spouse and parch is for parch
If passenger is traveling with spouse, sbsp is 1 and if they have children the parch 
will have number of children traveling with them.
If pasenger is child parch is either 0 or 1 or 2, if passenger is adult then parch is 
number of children

If passenger is traveling with sibling, sbsp is number of siblings and parch is number 
of parents 0 or 1 or 2 

So family size siblings/spouse plus parch plus the individual passenger count


```{r}
# Let's view family with size 8

big_family = data.combined %>% filter(parch == 9)
#View(big_family)

# The ticket number is CA. 2343, let's see all passengers with that ticket number

big_family_members = data.combined %>% filter(ticket == "CA. 2343")
#View(big_family_members)
big_family_members
```

Mr and Mrs Sage are traveling with 9 children, each child has 8 siblings 
so the family size is 9 children plus the couple that is 11

```{r}
str(data.combined$parch)

str(data.combined$sibsp)

#Since we have parch and sibsp are factors, we have to convert the levels
#to numeric to add

as.numeric(big_family_members$parch)
```

This returns the value of level assigned to the factor starting from number 0, 1 and
so on up to how many distinct values are there for that variable


```{r}
# Values of factors, can be converted to character then numeric
as.character(big_family_members$parch)
as.numeric(as.character(big_family_members$parch))

as.numeric(big_family_members$sibsp)
#This also returns the level assigned to the value of factor

# To get the actual value, get the character value then convert it to numeric 
as.character(big_family_members$sibsp)
as.numeric(as.character(big_family_members$sibsp))

#Let's apply for the entire dataset to get the family size
temp.parch = as.numeric(as.character(data.combined$parch))
head(temp.parch)
str(temp.parch)

temp.sibsp = as.numeric(as.character(data.combined$sibsp))
head(temp.sibsp)
str(temp.sibsp)

data.combined$family.size = as.factor(temp.parch + temp.sibsp + 1)
#View(head(data.combined))
head(data.combined)

big_family_members = data.combined %>% filter(ticket == "CA. 2343")
#View(big_family_members)
big_family_members

# Let's visualize to see if family.size has any predictive power
#x-axis family size which is factor so use bar plot broken by pclass and title

ggplot(data.combined[1:891,], aes(x = family.size, fill = survived)) +
  geom_bar() +
  facet_wrap(~ pclass + title) +
  ggtitle("Survival wrt family size broken by pclass and title") 

```



## Explore ticket feature

Let's explore the ticket variable.

Tickets were bought for group not individual tickets for each seat as in modern day
based on individuals with same ticket.
Fare can be calculated for each individual based on title and age


```{r}
# group by tickets and engineer below 13 features 

ticket.lookup = train %>%
  group_by(ticket) %>%
  summarize(group.count = n(),
            avg.fare = max(fare)/n(),
            female.count = sum(sex == "female"),
            male.count = sum(sex == "male"),
            child.count = sum(age < 18),
            elderly.count = sum(age > 54.0),
            female.ratio = sum(sex == "female")/n(),
            male.ratio = sum(sex == "male")/n(),
            child.ratio = sum(age < 18)/n(),
            elderly.ratio = sum(age > 54.0)/n(),
            female.child.ratio = ( sum(age <18) +
                                     sum(sex == "female" & age >= 18))/n(),
            min.age = min(age),
            max.age = max(age))



group_tickets = train %>% 
  group_by(ticket) %>% 
  summarise(n = n()) %>%
  filter( n > 1)

glimpse(group_tickets)
sum(group_tickets$n)

group_tickets %>% filter(ticket == "3101295")

#View(group_tickets %>% filter(ticket == "3101295"))
head(group_tickets %>% filter(ticket == "3101295"))


single_tickets = train %>% 
  group_by(ticket) %>% 
  summarise(n = n()) %>%
  filter( n == 1)
sum(single_tickets$n)

glimpse(single_tickets) 

#View(ticket.lookup)
head(ticket.lookup)

names(train)

#Double-check our work
ticket.lookup %>% filter(ticket == "3101295")

#View(ticket.lookup %>% filter(ticket == "3101295"))
head(ticket.lookup %>% filter(ticket == "3101295"))

#View(train %>% filter(ticket == "3101295"))
head(train %>% filter(ticket == "3101295"))


# Now lets add all the new features to the original df
train = train %>%
  left_join(ticket.lookup, by = "ticket")

#View(train %>% filter(ticket == "3101295"))
head(train %>% filter(ticket == "3101295"))

names(train)

```



### Analysing ticket variable for combined data without dplyr


```{r}
str(data.combined$ticket)

head(data.combined$ticket)

data.combined$ticket[1:20]
```

There's no immediately apparent structure in the data, let's see if we can find some.
We'll start with taking a look at just the first char for each. 

```{r}
# if empty ticket string replace with space o.w extract the first character
ticket.first.char = ifelse(data.combined$ticket == ""," ", 
                           substr(data.combined$ticket,1,1))

#View(ticket.first.char)
head(ticket.first.char)
sum(is.na(ticket.first.char))
length(unique(ticket.first.char))

#Let's add this as new feature and convert to factor
data.combined$ticket.first.char = as.factor(ticket.first.char)


# First high-level plot 
#x-axis ticket.first.char is factor so bar plot, fill with label survived

ggplot(data.combined[1:891,], aes(x = ticket.first.char, fill = survived)) +
  geom_bar() +
  ggtitle("Survival wrt first character of ticket")

#Looks like ticket number might be predictive, let's break down further wrt pclass
ggplot(data.combined[1:891,], aes(x = ticket.first.char, fill = survived)) +
  geom_bar() +
  facet_wrap(~pclass) +
  ylim(0,300) +
  ggtitle("Survival wrt first character of ticket and pclass") 
  
# Lastly, see if we get a pattern when using combination of pclass & title
ggplot(data.combined[1:891,], aes(x = ticket.first.char, fill = survived)) +
    geom_bar() +
    facet_wrap(~pclass+title) +
    ggtitle("Survival wrt first character of ticket, pclass and title") 


ggplot(data.combined[1:891,], aes(x = ticket.first.char, fill = survived)) +
    geom_bar() +
    facet_wrap(~pclass+title) +
    ylim(0,200) +
    ggtitle("Survival wrt first character of ticket, pclass and title") 
    
```


## Explore fares feature

Let's explore the fares variable to see if it has any predictive power to determine survival of passengers


```{r}
str(data.combined$fare)

summary(data.combined$fare)

# Visualization

# Can't make fare a factor, treat as numeric & visualize with histogram  
"
x-axis fare is numeric so histogram, let's just see the distribution for fare
"
ggplot(data.combined[1:891,], aes(x = fare))    +
  geom_histogram() +
  ggtitle("Survival wrt fare")

# Let's check to see if fare has predictive power   
ggplot(data.combined[1:891,], aes(x = fare, fill = survived))    +
  geom_histogram(binwidth = 5) +
  facet_wrap(~pclass + title) +
  ylim(0,50) +
  ggtitle("Survival wrt fare")   

```

### Imputate missing fares

What's with zero fares
How many passengers have zero fare? Get the subset 

```{r}
zero.fare = train %>% filter(fare == 0.0)

#View(zero.fare)
head(zero.fare)

dim(zero.fare)
```

15 out of 891 

They're all males traveling alone not specific to any class and they all
embarked in 'S'. All but one perished.
This is a classification problem to train model to accurately predict label Survival
If we only had this data or zero.fare, 14 out 15 perished, so if we predicted
that if fare is zero predict perished the model would be 93 % accurate

14/15*100

But this data is so skewed, all males, all embarked S.
So let's explore this data further

```{r}
# Let's get the totals by Pclass
# n() is function in dplyr, total number of observations in each group

zero.fare.pclass = zero.fare %>%
  group_by(pclass) %>%
  summarise(total = n()) %>%
  arrange(desc(total))
```

summarize - how many observations in each groups. 
How many observations in each Pclass

```{r}
zero.fare.pclass
```

So it is clear that zero.fare is not specific to any class
"
What is the distribution of fare for passengers similar to fare 0.0
so that we can replace the 0.0 value for example, replace 0 fare for male in class 3
with mean or median fare of males in class 3

zero fares are all males travelling alone so we can say they're all adult males

We can find the avg fare for adult males travelling alone

Most of the zero fares for males, that is title Mr.
So we use this correlation between title and fares to impute the zero fares

Now going back to zero fares impute zero fares 
we'll use replacement , you can also use a model to predict the fares
for males group by Pclass.
Beauty of dplyr is that the results can be stored as dataframe


Finding the summary of fares for males grouped by Pclass.
summarise is for aggregate functions
we will see which of the statistics we can use to impute fare

```{r}
mr.fare.stats = train %>%
  filter(title == "Mr.") %>%
  group_by(pclass) %>%
  summarise(fare.Min = min(fare),
            fare.Max = max(fare),
            fare.Median = median(fare),
            fare.Var = var(fare),
            fare.SD = sd(fare),
            fare.IQR = IQR(fare))


#View(mr.fare.stats)
head(mr.fare.stats)

str(mr.fare.stats)

#The above can be done as below
summary(train$fare[train$title == "Mr." & train$pclass == 1])
```

And you have to repeat for each Pclass,that for 2, 3 and it won't be in tabular 
format, so the above code is definitely more elegant because we can see the results 
in data frame

```{r}
#mean/median
11.29976/7.8958
```

1.43111
amost 1.5 times bigger

```{r}
11.29976/7.8958*100
```

mean is almost 43 percent more than median
fare is same for each member of family
so big families may have skewed the fare
so it makes sense to use median to impute fare


In order to not lose the original data 
so that if imputed values do not improve prediction accuracy
we can revert to original values,we use tracking features for that


Create "tracking feature" for those records that originally had 0.0 for the Fare variable. Common practise in Data Science to identify imputed value.

```{r}
train$fare.zero = ifelse(train$fare == 0.0, "Y", "N")

head(train)

table(train$fare.zero)
```

Impute the zero fares using the lookup table new column has median.

```{r}
#create the new df with median fare for mr. grouped by Pclass

zero.fare.lookup = train %>%
  filter(title == "Mr.") %>%
  group_by(pclass, title) %>%
  summarize(new.fare = median(fare))

#View(zero.fare.lookup)
head(zero.fare.lookup)

```

Now for each matching Mr. and Pclass add the New.Fare value
update zero Fare with New.Fare and then remove New.Fare

```{r}
train = train %>%
  left_join(zero.fare.lookup, by = c("pclass", "title")) %>%
  mutate(fare = ifelse(fare == 0.0, new.fare, fare)) %>%
  select(-new.fare)

#View(train)
head(train)

train.zero.fare = train %>%
  filter(fare == 0.0)

#View(train.zero.fare)
head(train.zero.fare)

train.zero.fare = train.zero.fare %>%
  select(fare)

# there are no zero fares
rm(train.zero.fare)

summary(train$fare)

```

## Explore cabin feature 

Let's explore the cabin variable to see if has any predictive power to determine survival of passengers

```{r}
str(data.combined$cabin)

sum(complete.cases(data.combined$cabin))  

head(data.combined$cabin)

#Replace the empty cabins with  "U"
data.combined = data.combined %>% mutate(cabin = ifelse(cabin == "","U", cabin))

head(data.combined$cabin)

# Take a look at the first char 
cabin.first.char = substr(data.combined$cabin,1,1)

str(cabin.first.char)

length(unique(cabin.first.char))

cabin.first.char = as.factor(cabin.first.char)

data.combined$cabin.first.char = cabin.first.char

names(data.combined)

head(data.combined$cabin.first.char)

# First high level plot
# x-axis cabin.first.char is a factor so bar plot

ggplot(data.combined[1:891,], aes(x = cabin.first.char, fill = survived)) +
  geom_bar() +
  ggtitle("Survical wrt first character of cabin")


ggplot(data.combined[1:891,], aes(x = cabin.first.char, fill = survived)) +
  geom_bar() +
  facet_wrap(~pclass) +
  ggtitle("Survical wrt first character of cabin, plcass and title")

# Does this feature improve upon pclass + title?
ggplot(data.combined[1:891,], aes(x = cabin.first.char, fill = survived)) +
  geom_bar() +
  facet_wrap(~pclass+title) +
  ggtitle("Survical wrt first character of cabin, plcass and title")


# What about folks with multiple cabins?
data.combined = data.combined %>% mutate(cabin.multiple = ifelse(
  str_detect(cabin," "), "Y", "N"))


head(data.combined$cabin.multiple)

names(data.combined)

length(unique(data.combined$cabin.multiple))

data.combined$cabin.multiple = as.factor(data.combined$cabin.multiple)

# x-axis cabin.multiple is factor so bar plot, broken by categorical variables
# pclass and title

ggplot(data.combined[1:891,], aes(x = cabin.multiple, fill = survived)) +
  geom_bar() +
  facet_wrap(~pclass + title) +
  ggtitle("Survical wrt multiple cabins or not, plcass and title")

```


## Explore embarked feature

```{r}
str(data.combined$embarked)

length(unique(data.combined$embarked))

table(data.combined$embarked)

data.combined$embarked = as.factor(data.combined$embarked)

# Does survivability depend on where you got onboard the Titanic?
#x-axis factor variable embarked so bar plot broken by categorical variables
#pclass and title 

ggplot(data.combined[1:891,], aes(x = embarked, fill = survived)) +
  geom_bar() +
  facet_wrap(~pclass + title) +
  ggtitle("Survivability wrt embarked, plcass and title")

```

# Visualization

The payoff
Investigate the hypothesis that traveling with children might be predictive

```{r}

# Set up factors
train$survived = as.factor(train$survived)
train$pclass = as.factor(train$pclass)


# Subset passengers traveling with children
tickets.children = train %>%
filter(child.count > 0)

```

## Visualize Survival rates for Ticket groups traveling with children

{show survival of children pclass-wise
  plot bar chart, fill by survived and faceted by Title
  Give title to your plot Survival Rates for Ticket Groups Traveling
  with children
  Lable y-axis as Count of Passengers}


```{r}

ggplot(tickets.children, aes(x=pclass, fill=survived))+
theme_bw() +
geom_bar() +
facet_wrap(~ title) +
labs(y = "Count of Passengers",
title = "Survival Rates for Ticket Groups Traveling with Children")

#or 

ggplot(tickets.children, aes(x=pclass, fill=survived))+
theme_bw() +
geom_bar() +
facet_wrap(~ title) +
ylab("Count of Passengers") +
ggtitle("Survival Rates for Ticket Groups Traveling with Children")
```

All boys in first, second class survived, in 3rd one-third survivied
almost same for Miss, Mrs
for comparison, no children
Subset passengers traveling with no children

```{r}
tickets.no.children = train %>%
filter(child.count == 0)

# Visualize
ggplot(tickets.no.children, aes(x=pclass, fill = survived)) +
theme_bw() +
geom_bar() +
facet_wrap(~ title) +
labs(y = "Count of Passengers",
title = "Survival Rates for Ticket Groups Traveling without Children")

#or 

ggplot(tickets.no.children, aes(x=pclass, fill = survived)) +
theme_bw() +
geom_bar() +
facet_wrap(~ title) +
ylab("Count of Passengers")+
ggtitle("Survival Rates for Ticket Groups Traveling without Children")
```

path to save the data.combined dataset with all the new features to be used for exploratory modeling

```{r eval=FALSE}
path
```

```{r eval=FALSE}
saveRDS(data.combined, paste0(path,"/RdsObjects/data.combined.rds"))
```









